---
title: "ukb_plasma_proteomics_analysis"
author: "Jonathan Chan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tictoc)
library(ggrepel)
library(ggsignif)
library(car)
library(readxl)
rm(list=ls())

theme_set(theme_classic())

date <- as.POSIXlt(Sys.time(), 'UTC')
date <- strftime(date, "%Y%m%dT%H:%M:%S%z")
date <- str_match(date,'^(\\d+)T')[,2]
```

#Olink Cardiometabolic Panel - UKB PP Extraction

This details some of the R code used in the extraction of data from the UKB PPP via Spark + JupyterLab in UKB RAP.

```{r ukbrap_pp_extract_prep}
olink_cm_panel <- read_csv2('DATA/UKB/RAP/Olink Explore Cardiometabolic - 2024-07-08.csv') %>%
  mutate(ukb_fieldname_equivalent = str_to_lower(Gene))

olink_cm2_panel <- read_csv2('DATA/UKB/RAP/Olink Explore Cardiometabolic II - 2024-07-08.csv') %>%
  mutate(ukb_fieldname_equivalent = str_to_lower(Gene))

olink_cm_panel <- bind_rows(olink_cm_panel, olink_cm2_panel)

#Fix nomenclature differences
olink_cm_panel$ukb_fieldname_equivalent[olink_cm_panel$Gene=='NT-proBNP'] <- 'ntprobnp'
olink_cm_panel$Gene[olink_cm_panel$Gene=='NT-proBNP'] <- 'NTproBNP'
olink_cm_panel$ukb_fieldname_equivalent[olink_cm_panel$Gene=='ERVV-1'] <- 'ervv_1'
olink_cm_panel$ukb_fieldname_equivalent[olink_cm_panel$Gene=='BOLA2'] <- 'bola2_bola2b'
olink_cm_panel$Gene[olink_cm_panel$Gene=='BOLA2'] <- 'BOLA2_BOLA2B'
rm(olink_cm2_panel)

# ukb_pp_fieldnames <- read.table('DATA/UKB/RAP/ukb_pp_fieldnames_240708.txt')
# print(str_c('Of the',nrow(olink_cm_panel),'OLink cardiometabolic panel, ',sum(olink_cm_panel$ukb_fieldname_equivalent %in% ukb_pp_fieldnames$V1), 'are found in the UKB PP',sep=' ')) #So all 369 CM panel found
# missing <- filter(olink_cm_panel, !ukb_fieldname_equivalent %in% ukb_pp_fieldnames$V1)
# rm(ukb_pp_fieldnames,missing)
# 
# #Generate the entity_table.field_name format for UKB
# entity_table <- c('olink_instance_0','olink_instance_2','olink_instance_3') #Instance 0 = main; 2= COVID-19 imaging and some imaging participants; 3=COVID-19 imaging participants
# cm_filenames <- c('eid', olink_cm_panel$ukb_fieldname_equivalent)
# entity_field <- map(entity_table, ~str_c(., cm_filenames, sep='.')) %>% unlist()
# 
# entity_field_tb <- tibble(entity_field)
# 
# write_tsv(entity_field_tb, 'DATA/UKB/RAP/ukb_pp_CMpanel_entityfield.tsv')
# rm(entity_field_tb)
```
Repeat for Olink's inflammation panel and other panels.

```{r}
sun2023_pp <- read_excel('DATA/UKB/RAP/Sun2023_UKBPP_PPs.xlsx') %>%
  select(Panel=`Protein panel`, Target=`Assay Target`, Gene=`Gene symbol`, UniProt)

cmpanel_sun23 <- sun2023_pp %>%
  filter(str_detect(Panel,'Cardiometabolic'))

length(cmpanel_sun23$Target[(arrange(cmpanel_sun23, Target)$Target != arrange(olink_cm_panel, Gene)$Gene)]) #Check for differences and no differences means all good

cmpanel_sun23 <- cmpanel_sun23 %>%
  left_join(select(olink_cm_panel, Gene, ukb_fieldname_equivalent), by=c('Target'='Gene'))
olink_cm_panel <- cmpanel_sun23 #Use the Sun 2023 ST3 as the base
rm(cmpanel_sun23)

include_nonCM_panels = T
if(isTRUE(include_nonCM_panels)){
  olink_nonCM_panels <- sun2023_pp %>%
    filter(!str_detect(Panel,'Cardiometabolic')) %>%
    mutate(ukb_fieldname_equivalent = str_to_lower(Target))
  
  #Fix nomenclature differences 
  olink_nonCM_panels$ukb_fieldname_equivalent[olink_nonCM_panels$Target=='HLA-DRA'] <- 'hla_dra'
  olink_nonCM_panels$ukb_fieldname_equivalent[olink_nonCM_panels$Target=='HLA-E'] <- 'hla_e'
  olink_nonCM_panels$ukb_fieldname_equivalent[olink_nonCM_panels$Target=='HLA-A'] <- 'hla_a' 
 
  # olink_nonCM_panels$ukb_fieldname_equivalent[!olink_nonCM_panels$ukb_fieldname_equivalent %in% colnames(ukb_proteomics_data)] #No differences
}
rm(sun2023_pp)
```

#Analysis of UK Biobank Plasma Proteomics Data

This R Markdown describes the analysis of the UK Biobank plasma proteomics data in the context of hypertrophic cardiomyopathy.

## Olink Cardiometabolic Panel in UKB PPP Analysis

```{r import_ukb_data}

#N.B Time instancing only applies to HCM cases. HF and T2D are assumed to have no time lag so i0 cases are used for those.

hf_eid <- read_csv('DATA/UKB/PROCESSED/1_I0/ukb_rap_HF_20240209_i0cases.tsv')
t2d_cases <- read_tsv('DATA/UKB/PROCESSED/1_I0/ukb_rap_T2D_20240209_i0cases.tsv')

#Define the instance here - either
## i0 = Only considering cases diagnosed prior to date of attendance at instance 0 
## 5Y = Only considering i0 cases OR cases diagnosed within 5 years of attendance OR died within 5 years of attendance and had death cause record mentioning disease
## Total = Considering ALL cases recorded in the UKB regardless of date of diagnosis

# instance <- '3_TOTAL'
# instance <- '1_I0'
instance <- '2_5Y'

# instance <- '4_NEGTOTAL' #Those who are never HCM or HF patients

if(instance =='1_I0'){
  #HCM Cases - UKB prior to Instance 0
  ukb_cases <- read_csv('DATA/UKB/PROCESSED/1_I0/ukb_rap_HCM_20240209_i0cases.tsv')
  #HF Cases - UKB prior to Instance 0
  
} else if (instance=='2_5Y'){
  #HCM Cases - UKB 5Y
  ukb_cases <- read_csv('DATA/UKB/PROCESSED/2_5Y/ukb_rap_HCM_20240209_5Ycases.tsv')

} else if (instance %in% c('3_TOTAL','4_NEGTOTAL')){
  #HCM Cases - All UKB
  ukb_cases <- read_csv('DATA/UKB/RAP/Cases/ukb_rap_HCM_20240209.csv') %>%
  dplyr::rename(ID=1)
}
```


```{r import_ukb_data}
#Import in the PP
pp_i0 <- read_tsv('DATA/UKB/RAP/PP/ukb_allpp_i0.tsv', col_types=rep('n',3006))%>% #All numeric columns
  select(-`...1`) %>%
  select(contains('instance_0')) %>% #It contains some instance 2 columns for some reason
  rename_with(~str_match(.,'\\.(.+)')[,2]) %>%
  filter(if_any(!'eid', ~!is.na(.))) %>%
  mutate(non_na_count = rowSums(!is.na(.))-1) %>% #-1 for eid is non-NA
  #filter(non_na_count > 2) %>%
  mutate(instance=0)

# pp_i2 <- read_tsv('DATA/UKB/RAP/ukb_allpp_i2.tsv')
# %>% select(contains('instance_2')) %>% #COVID-19 imaging study 
#   rename_with(~str_match(.,'\\.(.+)')[,2])%>%
#   filter(if_any(nppb:igfbp6, ~!is.na(.)))%>%
#   mutate(instance=2)
# 
# pp_i3 <- read_tsv('DATA/UKB/RAP/ukb_allpp_i3.tsv') %>% select(contains('instance_3')) %>% #COVID-19 repeat imaging study
#   rename_with(~str_match(.,'\\.(.+)')[,2])%>%
#   filter(if_any(nppb:igfbp6, ~!is.na(.)))%>%
#   mutate(instance=3)
# 
# #Prioritise usage of instance 0 where available
# pp_i2 <- filter(pp_i2, !eid %in% pp_i0$eid) 
# pp_i3 <- filter(pp_i3, !eid %in% pp_i0$eid)

#Looks like i2 and i3 participants all have measurements in i0 as well so using those for now
ukb_proteomics_data <- pp_i0
rm(pp_i0, pp_i2, pp_i3)
```


```{r import_ukb_data}
#Filter for only the plasma proteins of choice
if (isTRUE(include_nonCM_panels)){
  ukb_pp_data <- ukb_proteomics_data %>%
    select(eid, instance, olink_cm_panel$ukb_fieldname_equivalent, olink_nonCM_panels$ukb_fieldname_equivalent) 
  
  ukb_pp_list <- bind_rows(olink_cm_panel, olink_nonCM_panels)#Some pp shared across multiple panels btw
  write_tsv(ukb_pp_list, 'DATA/UKB/PROCESSED/ukb_pp_list_allpanel.tsv')
} else{
  ukb_pp_data<- ukb_proteomics_data %>%
    select(eid, instance, olink_cm_panel$ukb_fieldname_equivalent)
  
  ukb_pp_list <- olink_cm_panel
  write_tsv(ukb_pp_list, 'DATA/UKB/PROCESSED/ukb_pp_list_cmpanel.tsv')
}
rm(ukb_proteomics_data)
```


```{r import_ukb_data_covar}
#Import in basic covariates
ukb_20240209_snptest_covariates <- read_tsv('DATA/UKB/RAP/Cohorts/2024_02_09_AllUKB_Age_Sex_Height_Weight_10PCs.tsv') %>%
  dplyr::rename(ID=`Participant ID`) %>%
  filter(ID %in% ukb_pp_data$eid)

colnames(ukb_20240209_snptest_covariates)[2:3] <- c('age','sex')
colnames(ukb_20240209_snptest_covariates)[12:21] <- str_c('pc',rep(1:10))

#Filter for instance 0 of height/weight/BMI
ukb_20240209_snptest_covariates <- ukb_20240209_snptest_covariates %>%
  select(-matches('Instance [123]'))
colnames(ukb_20240209_snptest_covariates)[4:5] <- c('height', 'weight')
ukb_20240209_snptest_covariates <- ukb_20240209_snptest_covariates %>%
  mutate(bmi = weight/(height/ 100)^2)
```

Add in the additional covariates such as blood pressure; T2D status; smoking status.

```{r dbp}

ukb_20240209_bp_i0 <- read_tsv('DATA/UKB/RAP/Cohorts/2024_02_09_AllUKB_Age_Sex_Height_Weight_10PCs_Instance0BP_smoking.tsv') %>%
  select(ID=eid, contains(c('p4079','p94','p4080','p93','p6177', #Manual and annotated SBP/DBP readings - two measures per instance
                            'p20116')))# Smoking status

#If automated readings present, set manual readings (p93/94) to NA

#For DBP
ukb_20240209_bp_i0 <- ukb_20240209_bp_i0 %>%
    mutate(p94_i0_a0=case_when(!is.na(p4079_i0_a0)|!is.na(p4079_i0_a1) ~ NA, #If automated reading present, ignore the manual readings
                                 T ~p94_i0_a0 ),
              p94_i0_a1=case_when(!is.na(p4079_i0_a0)|!is.na(p4079_i0_a1) ~ NA, #If automated reading present, ignore the manual readings
                                 T ~p94_i0_a1))

dbp_vals <- rowMeans(select(ukb_20240209_bp_i0,contains(c('p4079','p94'))),na.rm=T)#This takes the mean of the blood pressure measurements - there are 2x readings per individual 


#For SBP
ukb_20240209_bp_i0 <- ukb_20240209_bp_i0 %>%
    mutate(p93_i0_a0=case_when(!is.na(p4080_i0_a0)|!is.na(p4080_i0_a1) ~ NA, #If automated reading present, ignore the manual readings
                                 T ~p93_i0_a0 ),
              p93_i0_a1=case_when(!is.na(p4080_i0_a0)|!is.na(p4080_i0_a1) ~ NA, #If automated reading present, ignore the manual readings
                                 T ~p93_i0_a1))
sbp_vals <- rowMeans(select(ukb_20240209_bp_i0,contains(c('p4080','p93'))),na.rm=T)

ukb_20240209_bp_i0 <- select(ukb_20240209_bp_i0 ,ID, p6177_i0, p20116_i0) %>%
  bind_cols(dbp=dbp_vals) %>%
  bind_cols(sbp=sbp_vals)

rm(dbp_vals, sbp_vals)

#Add adjustment for presence of blood pressure medication as per p6177
ukb_20240209_bp_i0  <- ukb_20240209_bp_i0  %>%
  mutate(dbp = case_when(str_detect(p6177_i0, 'Blood pressure medication')~dbp + 10, 
                         T~dbp),
         sbp = case_when(str_detect(p6177_i0, 'Blood pressure medication')~ sbp + 15, 
                         T~sbp)) %>%
  mutate(pp=sbp-dbp)

print(sum(ukb_20240209_bp_i0$pp<0, na.rm=T)) #i.e if sbp < dbp = red flag


#Add to ukb_20240209_snptest_covariates
ukb_20240209_snptest_covariates <- ukb_20240209_snptest_covariates %>%
  left_join(select(ukb_20240209_bp_i0 , ID, dbp, 
                   #sbp,pp, 
                   smoking=p20116_i0), by='ID') %>%
  mutate(smoking = ifelse(smoking =='Prefer not to answer', NA, smoking))

rm(ukb_20240209_bp_i0)


#Add T2D status (using i0 cases)
ukb_20240209_snptest_covariates <- ukb_20240209_snptest_covariates %>%
  mutate(t2d = ifelse(ID %in% t2d_cases[[1]], T, F))
rm(t2d_cases)

write_tsv(ukb_20240209_snptest_covariates, 'DATA/UKB/PROCESSED/20240209_ukb_age_sex_htwt_10pcs_dbp_t2d_smoking_cov.tsv')
```
Add in further covariates recommended by Rob Clarke and Mohsen e.g eGFR (glomerular filtration rate); socioeconomic Townsend deprivation index; and HDL/LDL cholesterol levels.

```{r}
rcmm_cov=T

if(isTRUE(rcmm_cov)){
  ukb_20240209_rcmm_cov <- read_tsv('DATA/UKB/RAP/Cohorts/2024_02_09_AllUKB_Age_Sex_Height_Weight_10PCs_Instance0BP_Smoking_HDL_LDL_Creatinine_Townsend.tsv')%>%
    select(ID='Participant ID','Age at recruitment', 'Sex',33:36)
  colnames(ukb_20240209_rcmm_cov) <- c('ID','age','sex','Townsend','creatinine_i0','hdl_i0','ldl_i0') #HDL/LDL are as per the blood biochemistry assays
  
  ukb_20240209_rcmm_cov <- ukb_20240209_rcmm_cov %>%
    mutate(creatinine_i0 = creatinine_i0 * (113.12 / 10^6) * 1000 / 10 ) %>% #Convert micromole/L to mg/dL for the CKI formula using the micromolar mass in g/micromole -> g/L -> mg/L -> mg/dL
    filter(!is.na(creatinine_i0)) %>% #Remove NA individuals for creatinine
    mutate(eGFR=case_when(sex=='Female' ~ 
                            142 * min(creatinine_i0/0.7, 1)^-0.241 * max(creatinine_i0/0.7,1)^-1.2 * 0.9938^age * 1.012, 
                          sex=='Male' ~
                            142*min(creatinine_i0/0.9,1)^-0.302 * max(creatinine_i0/0.9,1)^-1.2 * 0.9938^age * 1
    )) %>% 
    select(ID, Townsend, eGFR,hdl= hdl_i0, ldl=ldl_i0)
  
  write_tsv(ukb_20240209_rcmm_cov, 'DATA/UKB/PROCESSED/20240209_ukb_rcmm_cov.tsv')
}
```


Append all the covariates and the HCM disease status with the HCM pp data.

```{r}
#With just age/sex/10PCs as covariates
# ukb_pp_hcm_cov <- ukb_pp_data %>%
#   mutate(hcm = ifelse(eid %in% ukb_cases$eid, T, F)) %>%
#   left_join(ukb_20240209_snptest_covariates,by=c('eid'='ID')) %>%
#   select(eid, instance, hcm:bmi, everything()) %>%
#   relocate(bmi, .after='weight') %>% #52667 individuals
#   filter(if_all(age:pc10, ~!is.na(.))) %>%#Filter out individuals who have NA value for covariates = 52384 individuals
#   filter(if_any(nppb:npc2, ~!is.na(.))) #Filter out individuals who have no non-NA values for CM panel proteins = 52372 individuals
# 
# write_tsv(ukb_pp_hcm_cov, 'data/UKB/PROCESSED/ukb_cmpanelpp_hcm_covariates.tsv')

#With bp; T2D; smoking status covariates
ukb_pp_hcm_cov2 <- ukb_pp_data %>%
  mutate(hcm = ifelse(eid %in% ukb_cases$ID, T, F)) %>%
  left_join(ukb_20240209_snptest_covariates,by=c('eid'='ID')) %>%
  select(eid, instance, hcm:t2d, everything()) %>%
  relocate(bmi, .after='weight') %>% #53015 individuals
  filter(if_all(age:t2d, ~!is.na(.))) %>%  #Filter out individuals who have NA value for covariates = 52104 individuals
  filter(if_any(all_of(ukb_pp_list$ukb_fieldname_equivalent), ~!is.na(.))) %>% #Filter out individuals who have no non-NA values for all panel proteins = 52104 individuals
  select(where(~any(!is.na(.)))) #Filter for plasma proteins which actually have non-NA values i.e remove glipr1

  #Write out the pp + covariates + HCM label depending on the time instancing
  if(isTRUE(include_nonCM_panels)){
      write_tsv(ukb_pp_hcm_cov2, str_c('DATA/UKB/PROCESSED/', instance, '/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking.tsv'))
  } else{
      write_tsv(ukb_pp_hcm_cov2, str_c('DATA/UKB/PROCESSED/', instance, '/ukb_cmpanelpp_hcm_covariates_bp_t2d_smoking.tsv'))
  }

if (isTRUE(rcmm_cov)){
  
  ukb_pp_hcm_cov2 <- ukb_pp_hcm_cov2 %>%
  left_join(ukb_20240209_rcmm_cov,by=c('eid'='ID')) %>%
  select(eid, instance, hcm:t2d, Townsend:ldl, everything()) %>% #52092 individuals
  select(-hdl) %>%  #Ignore HDL for now due to loss of non-NA individuals
  filter(if_all(c(Townsend,eGFR,ldl), ~!is.na(.)))#Filter out individuals who have NA value for covariates = 49576 individuals

  if(isTRUE(include_nonCM_panels)){
    write_tsv(ukb_pp_hcm_cov2, str_c('DATA/UKB/PROCESSED/', instance, '/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv'))
  } else{
    write_tsv(ukb_pp_hcm_cov2, str_c('DATA/UKB/PROCESSED/', instance, '/ukb_cmpanelpp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv'))
  }
} 

sum(ukb_pp_hcm_cov2$hcm)
```

Reimport data here

```{r}

if(isTRUE(include_nonCM_panels)){
    ukb_pp_hcm_cov2 <- read_tsv(str_c('DATA/UKB/PROCESSED/', instance, '/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv'))
    
    ukb_pp_list <- read_tsv('DATA/UKB/PROCESSED/ukb_pp_list_allpanel.tsv')
  } else{
    ukb_pp_hcm_cov2 <- read_tsv(str_c('DATA/UKB/PROCESSED/', instance, '/ukb_cmpanelpp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv'))
    ukb_pp_list <- read_tsv('DATA/UKB/PROCESSED/ukb_pp_list_cmpanel.tsv')
  }
```

# QC of Time-Instancing

This compares the two subgroups of HCM patients that are divided by the time-instancing approach - specifically this approach splits the total HCM cases in UKB into

1. Those diagnosed by time of blood sample collection or <5 years after = 42 cases
2. Those diagnosed >5 years of blood sample collection = 58 cases

I don't filter out the HF cases btw.

```{r}
#HCM Cases - Total
total_ukb_hcm <- read_csv('DATA/UKB/RAP/Cases/ukb_rap_HCM_20240209.csv') %>%
  dplyr::rename(ID=1)

#HCM Cases - UKB 5Y
fiveyear_ukb_hcm <- read_tsv('DATA/UKB/PROCESSED/2_5Y/ukb_rap_HCM_20240209_5Ycases.tsv')

#Non-5Y cases i.e 'incident'
fiveyear_incident_ukb_hcm <- total_ukb_hcm %>%
  filter(!ID %in% fiveyear_ukb_hcm$ID)

#This is the formatted tibble with all covariates
og_tb <- read_tsv(str_c('DATA/UKB/PROCESSED/3_TOTAL/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv')) %>%  #
  select(-hcm)

#Filter
og_tb <- og_tb %>%
  mutate(hcm_instance = case_when(
    eid %in% fiveyear_ukb_hcm$ID ~ '<5Y Case',
    eid %in% fiveyear_incident_ukb_hcm$ID ~ '>5Y Case',
    T ~ 'Control'
  ))

print(table(og_tb$hcm_instance))
```

Compare across

1) Age
2) NTproBNP and HRC (raw)
3) PC1 and PC2 scores (PCA computed over cases + controls and over all 2.9k plasma proteins)
```{r}
#Run PC
# nonNA_across_allpp <- og_tb %>% select(aamdc:zpr1) %>% drop_na() #749 of the 49k individuals have non-NA values across
# pca_obj <- prcomp(nonNA_across_allpp, na.action='na.omit', center=T, scale=T)
# 
# new_pc_scores <- predict(pca_obj, newdata = og_tb %>% select(aamdc:zpr1))
# rm(new_pc_scores)
# # Extract only PC1, PC2, and PC3
# new_pc_scores_pc1to3 <- as_tibble(new_pc_scores[, 1:3])

```


```{r}
phenos_to_compare <- c('age', 'ntprobnp','hrc')

dist_plotter <- function(pheno, input_tb, output_folder='OUTPUT/UKB/PLOTS/0_EDA/3_HCM_SUBSET_COMP/'){
  
  medians <- input_tb %>%
    group_by(hcm_instance) %>%
    summarise(median_value = median(eval(parse(text=pheno)),na.rm=T), .groups = "drop")
  
  case_vs_case_wilcox_result<- wilcox.test(filter(input_tb,hcm_instance=='<5Y Case')[[pheno]], filter(input_tb, hcm_instance=='>5Y Case')[[pheno]], conf.int=T)
  
  p_value1 <- case_vs_case_wilcox_result$p.value
  conf_int1 <- case_vs_case_wilcox_result$conf.int
  
  # Print results for case vs. case
  print(paste("P-value (Prevalent vs. Incident Cases):", p_value1))
  print(paste("Confidence Interval:", conf_int1[1], "to", conf_int1[2]))
  
  incident_vs_control_wilcox_result<- wilcox.test(filter(input_tb,hcm_instance=='Control')[[pheno]], filter(input_tb, hcm_instance=='>5Y Case')[[pheno]], conf.int=T)
  
  p_value2 <- incident_vs_control_wilcox_result$p.value
  conf_int2 <- incident_vs_control_wilcox_result$conf.int
  
  # Print results
  print(paste("P-value (Incident vs. Controls):", p_value2))
  print(paste("Confidence Interval:", conf_int2[1], "to", conf_int2[2]))
  
  density_plot <- ggplot(input_tb, aes(x=eval(parse(text=pheno))))+
    geom_density(aes(fill=hcm_instance), alpha=0.5)+
    geom_vline(data = medians, aes(xintercept = median_value, color = hcm_instance), linetype = "dashed", linewidth = 0.5) +
    ylab('Density')+
    xlab(str_to_upper(pheno))+
    labs(caption = str_c('Wilcoxon Test between Case Subgroups: p-value = ', signif(p_value1,digits=3), ' and 95% CI = [',signif(conf_int1[1],digits=3), ',',signif(conf_int1[2],digits=3),']','\n Wilocoxon Test between >5Y Cases & Controls: p-value = ', signif(p_value2,digits=3), ' and 95% CI = [',signif(conf_int2[1],digits=3), ',',signif(conf_int2[2],digits=3),']'))+
    labs(fill='HCM Status', color='HCM Status')+
    scale_fill_brewer(palette='Set1')+
    scale_color_brewer(palette='Set1')
    
  print(density_plot)
  
  if(isFALSE(dir.exists(output_folder))){
    dir.create(output_folder)
  }
  
  ggsave(str_c(output_folder, pheno, '_hcm_subset_comp.png'),density_plot, dpi=600)
}

walk(phenos_to_compare, ~dist_plotter(., og_tb))

```




# Case-Control Evaluation of Plasma Proteins in UKB

This compares the levels of plasma proteins in HCM cases vs. controls in UKB.
This analyses the 736 proteins of the Olink Cardiometabolic I/II panels.

## QC

First we perform some QC.

```{r pp_qc}
#Visualise the distribution over all the proteins by plotting a scatter plot of mean and SD 

qc_p_tb <- ukb_pp_hcm_cov %>%
  select(nppb:npc2) %>%
  pivot_longer(everything(),names_to='pp',values_to='NPX') %>%
  group_by(pp) %>%
  summarise(mean=mean(NPX,na.rm=T), sd=sd(NPX,na.rm=T)) %>%
  unique()

qc_p_tb <- qc_p_tb %>%
  mutate(Label=case_when(abs(mean)>=0.5 | sd >= 2 ~ pp,
                         T~''))

qc_proteins <- ggplot(qc_p_tb, aes(x=sd, y=mean))+
  geom_point(alpha=0.5)+
  geom_vline(xintercept=0, linetype='dashed', col='red',alpha=0.3)+
  geom_hline(yintercept=0, linetype='dashed',col='red',alpha=0.3)+
  geom_text_repel(aes(label=Label))

print(qc_proteins)

#Visualise distribution for certain pp of interest

dist_visualiser <- function(input_tb, pp_of_interest,output_path='PLOTS/UKB/0_EDA/1_PP_DISTRIBUTIONS/'){
  
  plot_tb <- input_tb %>% select(eid,pp_of_interest)
  colnames(plot_tb) <- c('eid','pheno')
  
  density_plot <- ggplot(plot_tb,aes(x=pheno))+
    geom_density()+
    xlab('pp_of_interest')+
    labs(title=str_c('UKB Distribution for plasma protein ',pp_of_interest))
  
  print(density_plot)
  ggsave(str_c(output_path, pp_of_interest,'_dist.png'),dpi=600)
  
}

outlier_pps <- c('zp3','pm20d1','gh1','klk3')
walk(outlier_pps, ~dist_visualiser(ukb_pp_hcm_cov, .))

rm(qc_p_tb)

#Repeat after inverse-rank based normalisation
qc_p_tb2 <- ukb_pp_hcm_cov %>%
  select(nppb:npc2)

mean <- map_dbl(qc_p_tb2, ~.[!is.na(.)] %>% RNOmni::RankNorm() %>% mean(.,na.rm=T))
sd <- map_dbl(qc_p_tb2, ~.[!is.na(.)] %>% RNOmni::RankNorm() %>% sd(.,na.rm=T))

plot_tb <- tibble('pp'=colnames(qc_p_tb2),mean=mean,sd=sd) %>%
    mutate(Label=case_when(abs(mean)>=0.5 | sd >= 2 ~ pp,
                         T~''))

qc_proteins2 <- ggplot(plot_tb, aes(x=sd, y=mean))+
  geom_point(alpha=0.5)+
  geom_vline(xintercept=0, linetype='dashed', col='red',alpha=0.3)+
  geom_hline(yintercept=0, linetype='dashed',col='red',alpha=0.3)+
  geom_text_repel(aes(label=Label))

print(qc_proteins2)

rm(mean, sd,qc_p_tb2, plot_tb)
```
The second plot after normalisation shows that rank-based inverse normalisation succeeds in normalising the distributions.


```{r multivariate_assessor}
# multivariate_assessor <- function(input_tb, response_var, predictor_vars, covars=c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5'),cont_or_discrete_response='cont', model_return=F, summary_print=T, ivnormalise=T){
#   
#   formula <- str_c(response_var, '~',
#                    str_c(covars, collapse='+'),'+',
#                    str_c(predictor_vars,collapse='+'))
#   
#   input_tb <- filter(input_tb, !is.na(!!sym(response_var))) #Exclude rows with NA values in the response variable
#   
#   if(isTRUE(ivnormalise)){
#     input_tb[[response_var]] <- RNOmni::RankNorm(input_tb[[response_var]])
#   }
#   
#   if(cont_or_discrete_response=='cont'){
#     model <- lm(formula, data=input_tb)
#   } else if (cont_or_discrete_response=='discrete'){
#     model <- glm(formula,data=input_tb, family='binomial' )
#   }
#   
#    if(isTRUE(summary_print)){
#       print(str_c('Phenotype',response_var, 'tested for association with', str_c(predictor_vars, collapse=', '),sep=' '))
#       print(summary(model))
#    }
#   
#   #Check for multicollinearity (using variance inflation factor)
#   if(sum(vif(model)>5)>=1){ #If any of the coefficients have VIF greater than 5
#     print('Variance inflation factor >5 detected in the model suggests multicollinearity')
#     stop()
#   }
#   
#   pvals <- summary(model)$coefficients
#   adjusted_rsq <- summary(model)$adj.r.squared
#   
#   if(predictor_vars == 'hcm'){ #i.e a discrete predictor
#     predictor_vars <- 'hcmTRUE'
#   }
#   
#   output_tb <- as_tibble(pvals) %>%
#     bind_cols('Predictor'=rownames(pvals)) %>%
#     mutate('Pheno'=response_var) %>%
#     select(Pheno,Predictor, everything()) %>%
#     filter(Predictor %in% predictor_vars) 
#   colnames(output_tb) <- c('Pheno','Predictor','Estimate','SE','tvalue','pval')
#   
#   output_tb <- output_tb %>% 
#     mutate(upperSE = Estimate + SE,
#            lowerSE = Estimate - SE) %>%
#     select(-tvalue)
#   
#   if (isTRUE(model_return)){
#     return(model)
#   } else{
#     return(output_tb)
#   }
# }
```

```{r summary_plotter}

# summary_plotter <- function(input_tb,output_label, output_path='PLOTS/UKB/1_Case_vs_Control/1_MultiLM_Summary/', pval='bonferroni', fdr_threshold=0.05,confint=F, se=F,test='t-Test', xlabel='Average Difference in Phenotype (Case - Control)'){
#   
#   if (pval=='fdr'){
#     input_tb <- input_tb %>%
#       mutate(pval = p.adjust(pval, 'fdr'))
#     pval_threshold <- fdr_threshold
#   } else if(pval=='bonferroni'){
#     pval_threshold <- 0.05/nrow(input_tb)
#   }
#   
#   input_tb <- input_tb %>% #Only label the ones which reach significance
#     mutate(Label=ifelse(pval < pval_threshold,Label, ''),
#            Label2=ifelse(pval < pval_threshold,Label2, ''),
#            FDR_significant = ifelse(pval < pval_threshold, T, F))
#  
#   #Base plot
#   summary_cont_plot <- ggplot(input_tb, aes(x=Estimate, y=-log10(pval), col=FDR_significant))+
#     geom_vline(xintercept=0, linetype='dashed')+
#     geom_point()+
#     scale_colour_manual(name='MTC Significant',values=c('FALSE'='grey','TRUE'='red'))+
#     geom_text_repel(aes(label=Label))+
#     labs(col='Phenotype',shape='Predictor')+
#     scale_x_continuous(n.breaks=10)+
#     xlab(xlabel)+
#     ylab('-log10(p-value)')+
#     labs(title=str_wrap(str_c('Summary plot of ', test,' tests for ',length(unique(input_tb$Pheno)), ' phenotypes across ', length(unique(input_tb$Predictor)), ' predictors')))
#   
#   
#   #Adjust the plot depending on the multiple testing correction
#   if (pval=='bonferroni'){
#     
#     summary_cont_plot <- summary_cont_plot +
#       geom_hline(yintercept=-log10(0.05/nrow(input_tb)), linetype='dashed')+
#       labs(caption=str_c('Bonferroni-corrected p-value threshold = ',signif(0.05/nrow(input_tb),3)))
#     
#   } else if (pval =='fdr'){
#     summary_cont_plot <- summary_cont_plot +
#       geom_hline(yintercept=-log10(pval_threshold), linetype='dashed')+
#       ylab('-log10(Adjusted p-value)')+
#       labs(caption='FDR correction applied for multiple testing burden')
#   }
#   
#   if(isTRUE(confint)){
#     summary_cont_plot <- summary_cont_plot +
#       geom_errorbar(aes(xmin=lowerCI, xmax=upperCI),alpha=0.5)
#     
#     append <- '_confint'
#   } else if(isTRUE(se)){
#       summary_cont_plot <- summary_cont_plot +
#       geom_errorbar(aes(xmin=lowerSE, xmax=upperSE),alpha=0.5)
#     
#     append <- '_SE'
#     }
#     else{
#     append <- ''
#   }
#   
#   print(summary_cont_plot)
#   ggsave(str_c(output_path,test,'_',output_label,'_summary_plot_',pval,append,'.png'),summary_cont_plot,dpi=600, width=12, height=6)
#   
#   return(input_tb)
# }
```

Define the plasma proteins of interest.

```{r}
pp <- ukb_pp_list$ukb_fieldname_equivalent[ukb_pp_list$ukb_fieldname_equivalent %in% colnames(ukb_pp_hcm_cov2)] %>%  #In case any PP are removed due to lack of non-NA values
  unique()
```

## Multivariate Regression

This performs the comparison but with adjustment for confounders (age; sex; BMI; common genetic PCs to reflect ancestry/population structure).

```{r}
#By default, it performs rank-based inverse normalisation of the response variable
# pp_vs_hcm_marginal <- map(pp, ~multivariate_assessor(ukb_pp_hcm_cov,.,'hcm',covars=c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5'),cont_or_discrete_response='cont', model_return=F, summary_print=F, ivnormalise = T)) %>% bind_rows() %>%
#   mutate(Label=olink_cm_panel$Gene, Label2=olink_cm_panel$Gene)

#Summary plotting
# pp_vs_hcm_marginal_bonferroni_summarytb <- summary_plotter(pp_vs_hcm_marginal, 'pp_vs_hcm') 
# pp_vs_hcm_marginal_fdr_summarytb <- summary_plotter(pp_vs_hcm_marginal, 'pp_vs_hcm_fdr_5percent', pval = 'fdr') #5% FDR
# pp_vs_hcm_marginal_fdr_summarytb_onepercent <- summary_plotter(pp_vs_hcm_marginal, 'pp_vs_hcm_fdr_1percent', pval = 'fdr', fdr_threshold=0.01)
# 
# write_tsv(filter(pp_vs_hcm_marginal_bonferroni_summarytb,FDR_significant==T),'PLOTS/UKB/1_Case_vs_Control/1_MultiLM_Summary/pp_vs_hcm_marginal_bonferroni_sig.tsv')
# write_tsv(filter(pp_vs_hcm_marginal_fdr_summarytb,FDR_significant==T),'PLOTS/UKB/1_Case_vs_Control/1_MultiLM_Summary/pp_vs_hcm_marginal_fdr_sig5percent.tsv')
# write_tsv(filter(pp_vs_hcm_marginal_fdr_summarytb_onepercent,FDR_significant==T),'PLOTS/UKB/1_Case_vs_Control/1_MultiLM_Summary/pp_vs_hcm_marginal_fdr_sig1percent.tsv')
```
This provides a boxplot function to look at the most significanat ones.

```{r}
box_plotter <- function(input_tb, pheno, output_path='output/2_CrossPheno_AssocAnalysis/ukb/EDA/case_vs_controls/', grouping_var='hcm', xunits='NPX'){
  
  boxplot <- ggplot(input_tb, aes(y=eval(parse(text=pheno)), x=eval(parse(text=grouping_var))))+
    geom_signif(comparisons=list(c('TRUE','FALSE')), test='t.test')+
    geom_boxplot()+
    xlab(str_to_upper(grouping_var))+
    ylab(str_c(pheno , ' (',xunits,')'))+
    labs(title=str_wrap(str_c(pheno, 'for', sum(input_tb[[grouping_var]]==T),'HCM cases vs.', sum(input_tb[[grouping_var]]==F),'non-HCM controls in UK Biobank', sep=' '),width=60),
         caption='Statistical test = 2-tailed t-test')
  
  print(boxplot)
  ggsave(str_c(output_path, pheno,'.png'), dpi=600)
}

# multilm_sigpp <- pp_vs_hcm_marginal_bonferroni_summarytb$Pheno
# walk(multilm_sigpp, ~box_plotter(ukb_pp_hcm_cov,.))
```

## Limma

This uses  differential protein expression analysis in Limma to compare the UKB Cardiometabolic proteins across HCM cases and controls.

This is as per UKB RAP tutorial: https://github.com/dnanexus/UKB_RAP/blob/main/proteomics/protein_DE_analysis/2_differential_expression_analysis.ipynb but implemented in R instead of Python.

```{r limma}
library(limma)

limma_dpe <- function(input_tb, protein, predictor_vars='hcm', covar=c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking'), ivnormalise=T){
  
  input_tb <- filter(input_tb, !is.na(!!sym(protein))) #Exclude rows with NA values in the response variable
  
  if(isTRUE(ivnormalise)){
    input_tb[[protein]] <- RNOmni::RankNorm(input_tb[[protein]])
  }
  
  #Separate into pheno_df for covariates and predictors and npx_df containing the NPX value for the protein over all individuals
  pheno_df <- select(input_tb, all_of(predictor_vars),all_of(covar)) %>%
    mutate(across(all_of(predictor_vars), ~factor(.)))
  npx_df <- select(input_tb, !!sym(protein))
  
  #Need to define the formula i.e model and the model matrix
  formula <- str_c('~0 +',
                   str_c(predictor_vars,collapse='+'),
                   '+',str_c(covar, collapse='+'))
  design <- model.matrix(eval(parse(text=formula)), pheno_df)
  
  #Need to create the matrix array for limma i.e rows = proteins; columns = samples via t() for transpose
  #Perform the model fitting using weighted least-squares
  lin_model <- lmFit(t(npx_df), design)
  
  #To obtain log-fold changes between groups of the predictor_var i.e hcmTRUE vs. hcmFALSE, you need to obtain the contrast of this fitted linear model
  if(predictor_vars=='hcm'){
      contr <- makeContrasts(hcmTRUE - hcmFALSE, levels = design)
  } else if (predictor_vars=='hf'){
    contr <- makeContrasts(hfTRUE - hfFALSE, levels = design)
  }else if (predictor_vars=='rare'){
    contr <- makeContrasts(rareTRUE - rareFALSE, levels = design)
  }
  
  # Estimate contrast for the protein of interest
  tmp <- contrasts.fit(lin_model, contr)
  
  # Empirical Bayes smoothing of standard errors (shrinks standard errors that are much larger or smaller than those from other proteins towards the average standard error)
  tmp <- eBayes(tmp)
  
  #Return results for that particular protein
  top.table <- topTable(tmp, sort.by = "P", n = Inf)
  results <-  as_tibble(top.table) %>%
    select(-5) %>%
    mutate(pp=protein) %>%
    select(pp, everything())
  
  return(results)
  
}
```

Visualise the results in a volcano plot

```{r}
summary_plotter_limma <- function(input_tb, output_label='ukb_allpanel', predictor = 'HCM status', output_path=str_c('PLOTS/UKB/1_Case_vs_Control/2_Limma/',instance), pval='bonferroni', fdr_threshold = 0.05, test='Empirical Bayes moderated t-test',xlabel='Log2(Fold-Change)', xlimits=c(-1,2),ylimits=c(0,40), specific_label=F, max_repel_overlaps=100){
  
  if (pval=='fdr'){
    input_tb <- input_tb %>%
      mutate(P.Value = p.adjust(P.Value, 'fdr'))
    pval_threshold <- fdr_threshold
  } else if(pval=='bonferroni'){
    pval_threshold <- 0.05/nrow(input_tb)
  } else if (is.numeric(pval)){ #i.e if defining own pvalue threshold to use
    pval_threshold <- pval
  }
  
  if(isFALSE(specific_label)){
    input_tb <- input_tb %>% #Only label the ones which reach significance
      mutate(Label=ifelse(P.Value < pval_threshold,Gene, ''),
           MTC_significant = ifelse(P.Value < pval_threshold, T, F)) %>%
      arrange(Label)
  } else {
    input_tb <- input_tb %>%
      mutate(MTC_significant = ifelse(Label=='', F, T)) %>%#Just for downstream code purposes to label some specific points
      arrange(Label)
  }

 
  #Base plot
  summary_cont_plot <- ggplot(input_tb, aes(x=logFC, y=-log10(P.Value), col=MTC_significant))+
    geom_vline(xintercept=0, linetype='dashed')+
    geom_point()+
    scale_colour_manual(name=ifelse(isFALSE(specific_label),'MTC Significant','PP of Interest'),values=c('FALSE'='grey','TRUE'='red'))+
    geom_text_repel(aes(label=Label), nudge_y=3 ,
                    #nudge_x=0.5,
                    force =5, max.overlaps=max_repel_overlaps)+
    labs(col='Phenotype',shape='Predictor')+
    scale_x_continuous(n.breaks=10, limits=xlimits)+
    scale_y_continuous(limits=ylimits)+
    xlab(xlabel)+
    ylab('-log10(p-value)')+
    labs(title=str_c('Summary plot of ', test,' tests for ',length(unique(input_tb$pp)), ' plasma proteins for ', predictor, ' predictor variable'))
  
  
  #Adjust the plot depending on the multiple testing correction
  if (pval=='bonferroni'){
    
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(0.05/nrow(input_tb)), linetype='dashed')+
      labs(caption=str_c('Bonferroni-corrected p-value threshold = ',signif(0.05/nrow(input_tb),3)))
    
  } else if (pval =='fdr'){
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(pval_threshold), linetype='dashed')+
      ylab('-log10(Adjusted p-value)')+
      labs(caption=str_c( fdr_threshold*100, '% FDR correction applied for multiple testing burden'))
  } else if (is.numeric(pval)){
    summary_cont_plot <- summary_cont_plot +
      geom_hline(yintercept=-log10(pval_threshold), linetype='dashed')+
      labs(caption=str_c('Applied p-value threshold = ',signif(pval_threshold,3)))
  }
  
  print(summary_cont_plot)
  ggsave(str_c(output_path,output_label,'_summary_plot_',pval,'.png'),summary_cont_plot,dpi=600, width=9, height=6)
  
  #Also output a MAplot 
  ma_plot <- ggplot(input_tb, aes(x = log10(abs(AveExpr)), y = logFC)) +
    geom_point(aes(col= MTC_significant), alpha=0.5) +
    geom_text_repel(aes(label=Label))+
    labs(x = "Log10(Mean Normalised NPX)", y = xlabel) +
    scale_colour_manual(name='MTC Significant',values=c('FALSE'='grey','TRUE'='red'))+
    labs(title=str_c('MA plot for ', test, ' in UKB plasma proteins'))

  print(ma_plot)
  ggsave(str_c(output_path,output_label,'_ma_plot_',pval,'.png'),ma_plot,dpi=600, width=9, height=6)
  
  return(input_tb)

}

```

## HCM Case-Control

This is a redo of HCM case-control analysis but excluding all HF patients from both cases and controls.
So the first initial discovery was HF-agnostic i.e cases = (HCM cases, some with HF, some without) vs. controls (non-HCM but some with HF, most without).
Then the HF analysis was cases = (HF cases without HCM) vs. controls (non-HF & non-HCM).

So this is the for the most appropriate analysis you would keep the controls the same throughout all the comparisons so controls = non-HCM, non-HF controls and change the cases:

1. Non-HCM, HF cases (as above)
2. HCM, non-HF case (as below 2)
3. HCM cases (both non-HF and with HF) (as below 1)

```{r hcm_nonHF_nonHCM_controls}
#This is redoing the initial analysis but using non-HF and non-HCM as controls (as opposed to just using non-HCM) but using all HCM as cases
ukb_pp_hcm_cov <- ukb_pp_hcm_cov2  %>%
  mutate(hcm = case_when(eid %in% ukb_cases$ID ~ T, #HCM cases independent of HF status
                         !eid %in% ukb_cases$ID & !eid %in% hf_eid$ID ~ F, #Non-HF non-HCM controls
                         T~NA)) %>% #non-HCM and HF 
  filter(!is.na(hcm))

sum(ukb_pp_hcm_cov$hcm==T)
sum(ukb_pp_hcm_cov$hcm==F)

if (isTRUE(rcmm_cov)){ #If using the extended covariate list suggested by Rob Clarke and Mohsen
  covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking','Townsend','eGFR', 'ldl')
} else{
  covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking')
}

#Reduce the search space only to plasma proteins which have non-NA values in at least 1 HCM case e.g exlcude glipr1 and others
tic()
#Running by default with age/sex/genetic PCs/dbp/T2D/smoking status covariates but comparing HCM cases to non-HF/HCM controls
limma_hcm_2<- map(pp, ~limma_dpe(ukb_pp_hcm_cov,., covar=covars)) %>% bind_rows()
toc()

if (isTRUE(rcmm_cov)){
   extra_folders <- 'rcmm_cov/'
}
if(isTRUE(include_nonCM_panels)){
  extra_folders <- str_c(extra_folders, 'allpanels/')
} 
path <- str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/', extra_folders) #Define the folder for output of plots and tabular data
if(!dir.exists(path)){dir.create(path, recursive=T)}

limma_hcm <- left_join(limma_hcm, select(ukb_pp_list, Gene=Target, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent')) %>% unique()
write_tsv(limma_hcm, str_c(path,'limma_results.tsv'))

pp_vs_hcm_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(limma_hcm, 'allpp_vs_hcm',predictor = 'HCM status',output_path=path, xlimits=c(-1,1.75),ylimits=c(0,30)) 

#Also for 5%FDR and 1%FDR
pp_vs_hcm_marginal_fdr5_summarytb_limma <- summary_plotter_limma(limma_hcm, 'allpp_vs_hcm',predictor = 'HCM status',output_path=path, pval = 'fdr', xlimits=c(-1,1.75),ylimits=c(0,30)) 
pp_vs_hcm_marginal_fdr1_summarytb_limma <- summary_plotter_limma(limma_hcm, 'allpp_vs_hcm_fdr1',predictor = 'HCM status',output_path=path, pval = 'fdr', fdr_threshold=0.01, xlimits=c(-1,1.75),ylimits=c(0,30)) 

rm(ukb_pp_hcm_cov, path)
```


```{r hcm_wo_hf}
#Define your pp of interest to focus on
# pp_of_interest <- c('NTproBNP', 'NPPB','HRC','TNNI3','LTBP2','F7') #Instance 0 - 5%FDR
pp_of_interest <- c('NTproBNP', 'NPPB','HRC','TNNI3','LTBP2','F7','ANGPT2','ACE2','STC2') #5Y - 5% FDR
# pp_of_interest <- c('NTproBNP', 'NPPB','HRC','TNNI3','LTBP2','EDN1','ACE2','SHISA5','FABP3','F7','APOM') #OVERALL - Bonferroni

pp_of_interest <- filter(ukb_pp_list, Target %in% pp_of_interest)$ukb_fieldname_equivalent
```


```{r hcm_wo_hf}
ukb_pp_hcm_cov_exclHF <- ukb_pp_hcm_cov2%>%
  mutate(hcm = case_when(eid %in% ukb_cases$ID & !eid %in% hf_eid$ID ~ T, #Non-HF HCM cases
                         !eid %in% ukb_cases$ID & !eid %in% hf_eid$ID ~ F, #Non-HF non-HCM controls
                         T~NA)) #Either non-HCM and HF or both HF and HCM
  
ukb_pp_hcm_cov_exclHF <- filter(ukb_pp_hcm_cov_exclHF, !is.na(hcm)) #i.e remove all HF individuals

sum(ukb_pp_hcm_cov_exclHF$hcm==T) #Non-HF HCM cases
sum(ukb_pp_hcm_cov_exclHF$hcm==F) #Non-HF non-HCM controls

limma_hcm_exclHF <- map(pp, ~limma_dpe(ukb_pp_hcm_cov_exclHF,., covar=covars)) %>% bind_rows()

path <- str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/exclHFcases/', extra_folders) #Define the folder for output of plots and tabular data
if(!dir.exists(path)){dir.create(path, recursive=T)}

limma_hcm_exclHF <- left_join(limma_hcm_exclHF, select(ukb_pp_list, Gene=Target, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent'))
write_tsv(limma_hcm_exclHF, str_c(path,'limma_results.tsv'))

#Plot summary plots
pp_vs_hcm_exclHF_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(limma_hcm_exclHF, 'allpp_vs_hcm_exclHF',predictor = 'HCM status',output_path=path,xlimits=c(-1,1.75),ylimits=c(0,30)) 
pp_vs_hcm_exclHF_marginal_fdr5_summarytb_limma <- summary_plotter_limma(limma_hcm_exclHF, 'allpp_vs_hcm_exclHF',predictor = 'HCM status',output_path=path, pval='fdr', xlimits=c(-1,1.75),ylimits=c(0,30))

rm(ukb_pp_hcm_cov_exclHF, path)
```

```{r}
#Only HF-HCM patients (but only in Total instance due to otherwise insufficient sample size)

# if (instance =='3_TOTAL'){
#   ukb_pp_hcm_cov_onlyHF <- ukb_pp_hcm_cov2  %>%
#   mutate(hcm = case_when(eid %in% ukb_cases$ID & eid %in% hf_eid$ID ~ T, #HF and HCM cases
#                          !eid %in% ukb_cases$ID & !eid %in% hf_eid$ID ~ F, #Non-HF non-HCM controls
#                          T~NA)) #Either non-HCM and HF or both HF and HCM
#   
# ukb_pp_hcm_cov_onlyHF <- filter(ukb_pp_hcm_cov_onlyHF, !is.na(hcm)) #i.e remove all HF individuals
# 
# sum(ukb_pp_hcm_cov_onlyHF$hcm==T) #HF & HCM patients
# 
# #Rerun Limma
# limma_hcm_onlyHF <- map(pp, ~limma_dpe(ukb_pp_hcm_cov_onlyHF,.)) %>% bind_rows()
# 
# path <- str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/onlyHFcases/') #Define the folder for output of plots and tabular data
# write_tsv(limma_hcm_onlyHF, str_c(path,'cmpanel_limma_results.tsv'))
# 
# limma_hcm_onlyHF <- left_join(limma_hcm_onlyHF, select(olink_cm_panel, Gene, `Protein name`, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent'))
# pp_vs_hcm_onlyHF_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(limma_hcm_onlyHF, 'allpp_vs_hcm_onlyHF',predictor = 'HCM status',output_path=path,xlimits=c(-1,2)) 
# pp_vs_hcm_onlyHF_marginal_fdr5_summarytb_limma <- summary_plotter_limma(limma_hcm_onlyHF, 'allpp_vs_hcm_onlyHF',predictor = 'HCM status',output_path=path, pval='fdr',xlimits=c(-1,2)) 
# 
# rm(ukb_pp_hcm_cov_onlyHF,path)
# }

```

## HF Case-Control 

This repeats the case-control analysis for UKB individuals with diagnosis of heart failure (I50) but excluding those in the HCM patient/cases IDs.
The aim of this is to evaluate the specificity of the novel biomarkers detected.

'Cases' = HF patients excluding HCM patients
'Controls' = Non-HF and non-HCM individuals

```{r}
ukb_pp_hcm_cov_hf <- ukb_pp_hcm_cov2 %>%
  mutate(hf = case_when(eid %in% hf_eid$ID & !eid %in% ukb_cases$ID ~T,#Non-HCM HF cases
                        !eid %in% hf_eid$ID& !eid %in% ukb_cases$ID ~ F, #Non-HF non-HCM controls
                        T~NA)) #The NA values refer to the cases which are both HCM and HF (103 apparently in the PP) dataset

ukb_pp_hcm_cov_hf <- filter(ukb_pp_hcm_cov_hf,!is.na(hf))
print(str_c('There are', sum(ukb_pp_hcm_cov_hf$hf==T), 'non-HCM HF cases and', sum(ukb_pp_hcm_cov_hf$hf==F), 'non-HF, non-HCM controls in the UKB PP dataset', sep=' '))

hf_limma_results <- map(pp, ~limma_dpe(ukb_pp_hcm_cov_hf,.,predictor_vars='hf', covar=covars)) %>% bind_rows()


path <- str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HF/', extra_folders) #Define the folder for output of plots and tabular data
if(!dir.exists(path)){dir.create(path, recursive=T)}

hf_limma_results <- left_join(hf_limma_results, select(ukb_pp_list, Gene=Target, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent'))
write_tsv(hf_limma_results, str_c(path,'limma_results.tsv'))

#Plot total summary plot for HF
# summary_plotter_limma(hf_limma_results, 'allpp_vs_hf',predictor = 'HF status',output_path=path, xlimits=c(-1,1.75),ylimits=c(0,200) )
# summary_plotter_limma(hf_limma_results, 'allpp_vs_hf',predictor = 'HF status',output_path=path,pval='fdr', xlimits=c(-1,1.75),ylimits=c(0,200) )

#Plotting only on the pp_of_interest
hf_pp_vs_hcm_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(hf_limma_results %>% mutate(Label = ifelse(pp %in% pp_of_interest, Gene,'')), 'selectpp_vs_hf',predictor = 'HF status',output_path=path, specific_label=T, xlimits=c(-1,1.75),ylimits=c(0,200),max_repel_overlaps=10000) 
hf_pp_vs_hcm_marginal_fdr5_summarytb_limma <- summary_plotter_limma(hf_limma_results %>% mutate(Label = ifelse(pp %in% pp_of_interest, Gene,'')), 'selectpp_vs_hf',predictor = 'HF status',output_path=path,pval='fdr', specific_label=T, xlimits=c(-1,1.75),ylimits=c(0,200),max_repel_overlaps=10000 )

rm(ukb_pp_hcm_cov_hf, path)
```

# Gene Ontology/Pathway Analysis

This prepare the gene ID lists used for gene ontology analysis via UniProt Gene IDs as per the original `olink_cm_panel` variable.

```{r}
if(isTRUE(include_nonCM_panels)){
  write_tsv(select(ukb_pp_list, UniProt),'./DATA/UKB/RAP/olink_allpanel_uniprot.tsv', col_names=F) #Write out the background gene set
} else{
  write_tsv(select(ukb_pp_list, UniProt),'./DATA/UKB/RAP/olink_cmpanel_uniprot.tsv', col_names=F) #Write out the background gene set
}

write_tsv(select(filter(ukb_pp_list, Target %in% filter(pp_vs_hcm_marginal_bonferroni_summarytb_limma, MTC_significant==T)$Gene),UniProt),
          str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/',extra_folders,'bonferroni_pass_uniprotIDs.tsv'), col_names=F)

write_tsv(select(filter(ukb_pp_list, Target %in% filter(pp_vs_hcm_marginal_fdr1_summarytb_limma, MTC_significant==T)$Gene),UniProt),
          str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/',extra_folders,'fdr1_pass_uniprotIDs.tsv'), col_names=F)

write_tsv(select(filter(ukb_pp_list, Target %in% filter(pp_vs_hcm_marginal_fdr5_summarytb_limma, MTC_significant==T)$Gene),UniProt),
          str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/',extra_folders,'fdr5_pass_uniprotIDs.tsv'), col_names=F)

```

## ClusterProfiler
I also use FEA via ClusterProfiler from Bioconductor package.

```{r fea_clusterprofiler}
library(clusterProfiler)
library(org.Hs.eg.db)

limma_results <- read_tsv(str_c('OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/',extra_folders,'/limma_results.tsv'))
fdr5_pass <- read_tsv(str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM/',extra_folders,'fdr5_pass_uniprotIDs.tsv'), col_names=F)
background_uniprot <- read_tsv('DATA/UKB/RAP/olink_allpanel_uniprot.tsv',col_names=F) %>% unique() #For allpanels

```
### Gene Ontology Analysis
This runs both ORA for the differentially expressed proteins and GSEA for the entire list of proteins evaluated.

```{r}
#Run ORA

#Run Ontology analysis
ontologies <- c('BP','MF','CC')
ora_results <- map(ontologies, ~enrichGO(fdr5_pass$X1, 'org.Hs.eg.db',keyType = 'UNIPROT', ont=., universe=background_uniprot$X1, readable=T))
names(ora_results) <- ontologies

#Run KEGG analysis
ora_results[['KEGG']] <- enrichKEGG(fdr5_pass$X1, 'hsa',keyType = 'uniprot', universe=background_uniprot$X1)

#gseGO #Can also run GSEA on the entire list of proteins provided
```


```{r}
enrich_plotter<- function(input_enrichobj, output_path, output_name){
  
  print(head(attr(input_enrichobj, 'result'), 20)) #This accesses the head of the result attribute of the enrichResult object
  
  if(nrow(input_enrichobj) > 0){
      dt <- dotplot(input_enrichobj, showCategory=20, orderBy='p.adjust', decreasing=F ) + xlim(c(0,1))
      print(dt)
    #emap <- emapplot(input_enrichobj)
    
    ggsave(dt, filename=str_c(output_path, output_name, '_dot.png'), dpi=600, width=6, height=9)
    #gsave(emap, filename=str_c(output_path, output_name, '_emap.png'), dpi=600)
  } else{
    print(str_c('No significant enrichments for', output_name, 'after FDR 5% adjustment', sep=' ' ))
  }

}

path <- str_c('OUTPUT/UKB/PLOTS/3_GO/',instance,'/',extra_folders)

walk2(ora_results, c(ontologies, 'KEGG'), ~enrich_plotter(.x, path, .y))

```


# Cox Regression for Incident HCM Diagnosis

I also perform Cox regression to evaluate associations of plasma proteins with **incident** diagnosis of HCM in all of UKB-PPP (excluding cases of HCM at baseline or within instance 5Y).
N.B This is almost a separate script so run straight from here without prior ones

```{r import}
# rm(list=ls())
library(survival)

# instance <- '1_I0'
instance <- '2_5Y'
rcmm_cov<- T
include_nonCM_panels=T

# pp_scaling <- 'scale' #This flag is used if you want to flag the scales via scale()
pp_scaling <- 'ranknorm'

ukb_cases_total <- read_tsv('DATA/UKB/RAP/Cases/ukb_rap_HCM_20240209.csv')

if(instance=='1_I0'){
  ukb_cases <- read_tsv('DATA/UKB/PROCESSED/1_I0/ukb_rap_HCM_20240209_i0cases.tsv')
} else if (instance=='2_5Y'){
  ukb_cases <- read_tsv('DATA/UKB/PROCESSED/2_5Y/ukb_rap_HCM_20240209_5Ycases.tsv')
}

if(isTRUE(include_nonCM_panels)){
  ukb_pp <- read_tsv(str_c('DATA/UKB/PROCESSED/',instance,'/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv')) %>%  #
  select(-hcm)
  
} else{
  ukb_pp <- read_tsv(str_c('DATA/UKB/PROCESSED/',instance,'/ukb_cmpanelpp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv')) %>%  
  select(-hcm)
}

if (pp_scaling == 'scale'){
  pp <- ukb_pp %>% select('nppb':'npc2') %>%
    scale()
  
  ukb_pp <- ukb_pp %>%
    select(-('nppb':'npc2')) %>%
    bind_cols(pp)
  rm(pp)
  
}else if (pp_scaling == 'ranknorm'){
  ukb_pp <- ukb_pp %>%
    mutate(across(nppb:npc2, ~ replace(., !is.na(.), RNOmni::RankNorm(.[!is.na(.)]))))
}

if (isTRUE(rcmm_cov)){
   extra_folders <- 'rcmm_cov/'
}
if(isTRUE(include_nonCM_panels)){
  extra_folders <- str_c(extra_folders, 'allpanels/')
} 

hcm_tb2 <- readRDS('DATA/UKB/PROCESSED/3_TOTAL/ukb_diagnosis_dates_minimum.rds') #Min_diag_date2 refers to earliest diagnosis date of HCM from either self-reported; hospital or death cause record

ukb_outcomes <- read_tsv('../../HCM_Severity/DATASETS/UKB/RAP/Outcomes/Processed/allukb_outcome_ages.tsv')
```


```{r}
ukb_diag_cox_tb <- ukb_pp %>%
  left_join(select(ukb_outcomes,ID, lost_age, death_age, age_attend_i0, dob_approx), by=c('eid'='ID')) %>%
  left_join(select(hcm_tb2, ID,min_diag_date2), by=c('eid'='ID'))

ukb_diag_cox_tb <- ukb_diag_cox_tb %>%
  filter(!eid %in% ukb_cases$ID) %>% #Filter out baseline i0 HCM cases or 5Y cases
  mutate(datasetversion_age = time_length(ymd('2024-02-09') - dob_approx, unit='year')) %>%
  mutate(incidenthcm_status = ifelse (eid %in% ukb_cases_total[['0']], T,F)) %>% #Add in status
  rowwise() %>%
  mutate(incidenthcm_age = (ifelse(incidenthcm_status==T, time_length(min_diag_date2-dob_approx, unit='year'), min(datasetversion_age, death_age, lost_age, na.rm=T)))) %>% #Add in incident_hcm_age = if status, that will be min_diag_date2 and if not, then it will be the age at dataset version or lost to followup
  ungroup()

#write_tsv(ukb_diag_cox_tb, str_c('DATA/UKB/PROCESSED/',instance,'/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov_HCMdiagCox_pp', pp_scaling, '.tsv'))
```

## Multivariate Cox

This runs Cox regression with the covariates of interest, running marginal models over all the pp.

```{r}
multivariate_cox <- function(input_tb, marginal_predictor, covariates, status, outcome_var, outcome_levels, print_summary=T,return_predictor_results_only=T, return_model=F){ #Return only the coefficient and p-values for the marginal_predictor
  
  x <- str_c(c(covariates,marginal_predictor), collapse='+')
  y <- str_c('Surv(',outcome_var,',',status,')')
  formula <- as.formula(paste(y,'~',x))
  
  multivariate_cox <- coxph(formula, data=input_tb)
  if(isTRUE(print_summary)){
    print(summary(multivariate_cox))
  }
  
  x<- summary(multivariate_cox)
  
  if(isTRUE(return_model)){
    return(multivariate_cox)
  }
  
  if(isFALSE(return_predictor_results_only)){
  
  p.value<-signif(x$coef[1:nrow(x$coef),5], digits=3)
  beta<-signif(x$coef[1:nrow(x$coef),1], digits=3);#coeficient beta
  HR <-signif(x$coef[1:nrow(x$coef),2], digits=3);#exp(beta) = hazard ratio
  HR.confint.lower <- signif(x$conf.int[,"lower .95"], 3)
  HR.confint.upper <- signif(x$conf.int[,"upper .95"],3)
  se_beta <- signif(x$coef[1:nrow(x$coef),3], digits=3)
  vars <- rownames(x$coef)
    
  } else{
      p.value<-signif(x$coef[nrow(x$coef),5], digits=3)
  beta<-signif(x$coef[nrow(x$coef),1], digits=3);#coeficient beta
  HR <-signif(x$coef[nrow(x$coef),2], digits=3);#exp(beta) = hazard ratio
  HR.confint.lower <- signif(x$conf.int[,"lower .95"][nrow(x$coef)], 3)
  HR.confint.upper <- signif(x$conf.int[,"upper .95"][nrow(x$coef)],3)
  se_beta <- signif(x$coef[nrow(x$coef),3], digits=3)
  vars <- rownames(x$coef)[nrow(x$coef)]

  }
  
  res<-bind_cols(vars,beta, se_beta, HR, HR.confint.lower, HR.confint.upper, p.value) %>%
    mutate(predictor=str_to_upper(vars)) %>%
    mutate(outcome = factor(
      str_to_upper(str_match(outcome_var,'(.+)_age')[,2]),levels=outcome_levels, ordered=T
    ))
  
  names(res)<-c('var',"beta",'se_beta', "HR", "Lower_95CI_HR","Upper_95CI_HR", 
            "p.value", 'predictor','outcome')
  
  return(res)
}

```

```{r}
#Run marginal Cox regression models for each plasma protein with each outcome of interest
#pp <- c('ntprobnp','hrc')
if(isTRUE(rcmm_cov)){
  covariates <- c('sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','smoking','t2d', 'Townsend', 'eGFR', 'ldl')
} else{
  covariates <- c('sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','smoking','t2d')
}

multivariate_cox_results <- c(
  map(pp, ~multivariate_cox(ukb_diag_cox_tb,.,covariates,'incidenthcm_status','incidenthcm_age', outcome_levels=c('INCIDENTHCM'), print_summary = F))
) %>%
  bind_rows() %>%
  filter(!is.na(p.value))

multivariate_cox_results <- multivariate_cox_results %>%
  left_join(dplyr::select(ukb_pp_list,ukb_fieldname_equivalent,Target), by=c('var'='ukb_fieldname_equivalent')) %>%
  unique() %>%
  dplyr::rename(Label=Target)

saveRDS(multivariate_cox_results, str_c('OUTPUT/UKB/PLOTS/5_CoxHCMDiag/',instance,'/',extra_folders,'/allcmpp_mvCox_results_pp',pp_scaling, '.rds'))


# multivariate_cox_results <- readRDS(str_c('OUTPUT/UKB/PLOTS/5_CoxHCMDiag/',instance,'/allcmpp_mvCox_results.rds'))
```

```{r}
multivariate_coxph_plotter <- function(multivariate_results, plot_output_path, output_name, pp_scaling, FDR=5, specific_label=F, ylimits=''){
  
  if (!is.numeric(FDR)){ #Apply Bonferroni
    pval_threshold = 0.05/nrow(multivariate_results)
    mtc <- 'Bonferroni'
  } else{
    multivariate_results <- mutate(multivariate_results, p.value=p.adjust(p.value, method='BH'))
    pval_threshold <- FDR/100
    mtc <- str_c('FDR ',FDR, '%',sep='')
  }
  
  if(isFALSE(specific_label)){
    multivariate_results <- multivariate_results %>% #Only label the ones which reach significance
      mutate(Label=ifelse(p.value < pval_threshold,Label, ''),
           MTC_significant = ifelse(p.value < pval_threshold, T, F)) %>%
      arrange(Label)
  } else {
    multivariate_results <- multivariate_results %>%
      mutate(MTC_significant = ifelse(Label=='', F, T)) %>%#Just for downstream code purposes to label some specific points
      arrange(Label)
  }
  
  multivariate_plot <- ggplot(multivariate_results, aes(HR,-log10(p.value), col=MTC_significant))+
    geom_vline(xintercept=1, linetype='dashed',alpha=0.5)+
    geom_hline(yintercept=-log10(pval_threshold), linetype='dashed', alpha=0.5)+
    geom_point(alpha=0.75)+
    scale_colour_manual(name=ifelse(isFALSE(specific_label),'MTC Significant','PP of Interest'),values=c('FALSE'='grey','TRUE'='red'))+
    geom_text_repel(aes(label=Label), nudge_y=3, force = 5, max.overlaps=40)+
    #geom_errorbarh(aes(xmin=Lower_95CI_HR, xmax=Upper_95CI_HR,col=predictor),alpha=0.75, height=0.1)+
    xlab(ifelse(isFALSE(pp_scaling),'Hazard Ratio', 'Hazard Ratio per SD'))+
    ylab(ifelse(mtc=='Bonferroni','-log10 (Wald-test p value)', '-log10 (Adjusted Wald-test p value)'))+
    scale_x_continuous(n.breaks=10)+
    labs(title=str_wrap(str_c('Multivariate Cox Regression analysis for ', nrow(multivariate_results), ' plasma protein predictors')),caption=str_c(mtc, ' p-value Threshold Indicated'))+
    theme(legend.position="right")
  
  if(!identical(ylimits,'')){
    multivariate_plot <- multivariate_plot+
      scale_y_continuous(limits=ylimits)
  }
  
  print(multivariate_plot)
  ggsave(str_c(plot_output_path, 'cox_multivariate_', output_name,'.png'), dpi=600, width=9, height=6)
  
  return(multivariate_results)
}
```


```{r}
path <- str_c('./OUTPUT/UKB/PLOTS/5_CoxHCMDiag/', instance,'/',extra_folders,'/')

mvcox_incidentHCM_fdr5 <- multivariate_coxph_plotter(multivariate_cox_results, path, str_c('allCMpp_incidentHCMdiag_fdr5_pp',pp_scaling), pp_scaling=pp_scaling)

multivariate_coxph_plotter(multivariate_cox_results, path, str_c('allCMpp_incidentHCMdiag_fdr5_pp_ylim30',pp_scaling), pp_scaling=pp_scaling, ylimits=c(0,30))

multivariate_coxph_plotter(multivariate_cox_results, path, str_c('allCMpp_incidentHCMdiag_fdr1_pp',pp_scaling), FDR=1, pp_scaling=pp_scaling)
multivariate_coxph_plotter(multivariate_cox_results, path, str_c('allCMpp_incidentHCMdiag_bonferroni_pp', pp_scaling), FDR=F, pp_scaling=pp_scaling)

#Write out 
write_tsv(dplyr::select(filter(ukb_pp_list, ukb_fieldname_equivalent %in% filter(mvcox_incidentHCM_fdr5, MTC_significant==T)$var),UniProt),
          str_c(path,'fdr5_pass_uniprotIDs_pp',pp_scaling,'.tsv'), col_names=F)

rm(path,gap)
```

## Kaplan-Meier Analysis

This compares the top 25% vs. the bottom 75% using Kaplan Meier using the target variable of incident HCM diagnosis.

```{r}
#Split the top 25% and bottom 75%

contvar_binner <- function(input_tb, variable, threshold, signif=F){ #Assumes split into two groups
  newcolname <- str_c(variable,'_categorical')
  
  output <- mutate(input_tb,
                   temp=ifelse(eval(parse(text=variable)) >= threshold, str_c('>= ', threshold), str_c('< ', threshold)))
  colnames(output)[ncol(output)] <- newcolname
  
  return(output)
}

ukb_diag_cox_tb <- contvar_binner(ukb_diag_cox_tb, 'ntprobnp', signif(quantile(ukb_diag_cox_tb$ntprobnp,probs=0.75,na.rm=T),3)) %>%
  contvar_binner('hrc',signif(quantile(ukb_diag_cox_tb$hrc,probs=0.75,na.rm=T),3))
```


```{r}
km_survival_curv_plotter <- function(input_tb, categorical_var, status, outcome_var, plot_output_folder){ #outcome_var = Define the column for age at first occurence of selected outcome of interest
  
  print(str_c('Plotting Kaplan Meier survival curves for', deparse(substitute(categorical_var)),sep=' '))
    
    plot_types <- c('event')
    
    for (type in plot_types){
    
      params <- list(fun.time = substitute(outcome_var),
    fun.event = substitute(status),
    grouping = substitute(categorical_var), 
    fun.dat = substitute(input_tb),
    type=type)
          
          plot <- substitute(ggsurvplot(
      fit= survfit(Surv(time = fun.time, event = fun.event) ~ grouping, data = fun.dat),
      fun=type,
      pval=T, conf.int=T, risk.table='abs_pct',risk.table.col='strata',linetype='strata', surv.median.line='hv',xlab='Age',ncensor.plot=T,ylim=c(0,0.025)), params)
    
    plot <- eval.parent(plot)
  
    print(plot)
    
    if (isFALSE(dir.exists(plot_output_folder))){
      dir.create(plot_output_folder, recursive=T)
    }
    
    pdf(str_c(plot_output_folder,deparse(substitute(categorical_var)),'_kaplan_',type,'.pdf'), width=12, height=9)
    print(plot, newpage=F)
    dev.off()
    }
}
```


```{r kaplan_meier, fig.width=9, fig.height=10, message=F, warning=F}
outcomes_of_interest <- c('incidenthcm_age')

if(isTRUE(rcmm_cov)){
  plotpaths <- str_c('./OUTPUT/UKB/PLOTS/5_CoxHCMDiag/',instance,'/KM/', outcomes_of_interest,'/rcmm_cov/') #Need to run separate results because 103 cases -> 100 cases when using expanded covariates
} else{
  plotpaths <- str_c('./OUTPUT/UKB/PLOTS/5_CoxHCMDiag/',instance,'/KM/', outcomes_of_interest,'/') 
}


#Run for UKB - Overall Composite
km_survival_curv_plotter(ukb_diag_cox_tb,sex,incidenthcm_status,incidenthcm_age, plotpaths[1]) #Using sex
km_survival_curv_plotter(ukb_diag_cox_tb,ntprobnp_categorical,incidenthcm_status,incidenthcm_age, plotpaths[1]) #Using sex #Using top 25% vs. bottom 75% of NTproBNP
km_survival_curv_plotter(ukb_diag_cox_tb,hrc_categorical,incidenthcm_status,incidenthcm_age, plotpaths[1]) #Using sex #Using top 25% vs. bottom 75% of HRC

```




# Dataset Preparation for ML

This outlines some dataset preparation for ML purposes such as addition of cv_prs to the covariate dataset.

```{r}
instance <- '2_5Y'

ukb_pp <- read_tsv(str_c('DATA/UKB/PROCESSED/',instance,'/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv'))
ukb_cv_prs <- bind_rows(
  read_tsv('../../HCM_IntScore/ML/1_input/hcmr_ukb_agesex_train_imputedPRS_rareGRSdosage_covariates_cvprs.tsv')%>% mutate(ID = as.character(ID)),
  read_tsv('../../HCM_IntScore/ML/1_input/hcmr_ukb_agesex_validation_imputedPRS_rareGRSdosage_covariates_cvprs.tsv')%>% mutate(ID = as.character(ID)),
  read_tsv('../../HCM_IntScore/ML/1_input/ukb_test_imputedPRS_rareGRSdosage_covariates_cvprs.tsv') %>% mutate(ID = as.character(ID))
) %>% 
  filter(!str_detect(ID, 'HCR')) %>% 
   select(ID, cv_prs, PC1:PC10) %>% 
  mutate(ID = as.numeric(ID))

ukb_pp <- ukb_pp %>%
  left_join(ukb_cv_prs, by=c('eid'='ID'))
```

# Re-analysis on Training Set Only

This reruns the case-control association analysis only on the training set from ML (excluding the test set individuals).
This starts from the ukb_pp_hcm_cov2 tibble btw which is saved as .tsv.

```{r}
train_eids <- read_csv('OUTPUT/UKB/ML/1_data/1_hcm_cc_noprs/X_train_eid.csv')$eid
test_eids <- read_csv('OUTPUT/UKB/ML/1_data/1_hcm_cc_noprs/X_test_eid.csv')$eid

#This uses the 5Y time instancing of HF cases 
ukb_cases <- read_tsv('DATA/UKB/PROCESSED/2_5Y/ukb_rap_HCM_20240209_5Ycases.tsv')
colnames(ukb_cases) <- c('ID')

rcmm_cov<-T
include_nonCM_panels <- T

instance <- '2_5Y'
    
if(isTRUE(include_nonCM_panels)){
    ukb_pp_hcm_cov2 <- read_tsv(str_c('DATA/UKB/PROCESSED/', instance, '/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv'))
    ukb_pp_list <- read_tsv('DATA/UKB/PROCESSED/ukb_pp_list_allpanel.tsv')

  } else{
    ukb_pp_hcm_cov2 <- read_tsv(str_c('DATA/UKB/PROCESSED/', instance, '/ukb_cmpanelpp_hcm_covariates_bp_t2d_smoking_rcmmcov.tsv'))
    ukb_pp_list <- read_tsv('DATA/UKB/PROCESSED/ukb_pp_list_cmpanel.tsv')
  }

pp <- ukb_pp_list$ukb_fieldname_equivalent[ukb_pp_list$ukb_fieldname_equivalent %in% colnames(ukb_pp_hcm_cov2)] %>%  #In case any PP are removed due to lack of non-NA values
  unique()

```

```{r}
#This is using non-HF and non-HCM as controls (as opposed to just using non-HCM) but using all HCM as cases
ukb_pp_hcm_cov <- ukb_pp_hcm_cov2  %>%
  mutate(hcm = case_when(eid %in% ukb_cases$ID ~ T, #HCM cases independent of HF status
                         !eid %in% ukb_cases$ID & !eid %in% hf_eid$ID ~ F, #Non-HF non-HCM controls
                         T~NA)) %>% #non-HCM and HF 
  filter(!is.na(hcm))

#Filter for training set individuals here

ukb_pp_hcm_cov_train <- ukb_pp_hcm_cov %>%
  filter(eid %in% train_eids)

sum(ukb_pp_hcm_cov_train$hcm==T)
sum(ukb_pp_hcm_cov_train$hcm==F)

if (isTRUE(rcmm_cov)){ #If using the extended covariate list suggested by Rob Clarke and Mohsen
  covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking','Townsend','eGFR', 'ldl')
} else{
  covars <- c('age','sex','bmi','pc1','pc2','pc3','pc4','pc5','dbp','t2d','smoking')
}

#Reduce the search space only to plasma proteins which have non-NA values in at least 1 HCM case e.g exlcude glipr1 and others
tic()
#Running by default with age/sex/genetic PCs/dbp/T2D/smoking status covariates but comparing HCM cases to non-HF/HCM controls
limma_hcm_train<- map(pp, ~limma_dpe(ukb_pp_hcm_cov_train,., covar=covars)) %>% bind_rows()
toc()

if (isTRUE(rcmm_cov)){
   extra_folders <- 'rcmm_cov/'
}
if(isTRUE(include_nonCM_panels)){
  extra_folders <- str_c(extra_folders, 'allpanels/')
} 
path <- str_c('./OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/',instance,'/HCM_Train/', extra_folders) #Define the folder for output of plots and tabular data
if(!dir.exists(path)){dir.create(path, recursive=T)}
limma_hcm_train <- left_join(limma_hcm_train, select(ukb_pp_list, Gene=Target, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent')) %>% unique()
write_tsv(limma_hcm_train,str_c(path,'limma_results.tsv'))

pp_vs_hcm_train_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(limma_hcm_train, 'allpp_vs_hcm_train',predictor = 'HCM status',output_path=path, xlimits=c(-1,1.75), ylimits=c(0,35)) 

#Also for 5%FDR and 1%FDR
pp_vs_hcm_train_marginal_fdr5_summarytb_limma <- summary_plotter_limma(limma_hcm_train, 'allpp_vs_hcm_train',predictor = 'HCM status',output_path=path, pval = 'fdr', xlimits=c(-1,1.75), ylimits=c(0,35)) 

rm(ukb_pp_hcm_cov, path)
```




```{r}
sessionInfo()
```


# Archive

Previous association analyses between the ~2.9k plasma proteins assayed via Olink Explore 3072 and numerous binary/quantitative traits have been performed by Eldjarn et al, 2023. They do perform analysis for ICD code 42 i.e 'cardiomyopathy'.

I repeat analyses with a more specific focus on hypertrophic cardiomyopathy.

```{r import_data}

eldjarn23_sig_associations_i42_ukb <- read_csv('DATA/UKB/Eldjarn23_I42_Cardiomyopathy_Significant_Associations_UKB.csv') #The BI refers to British Individuals and AF refers to British/Irish ancestry; AF refers to African ancestries; SA refers to to South Asian ancestries
eldjarn23_sig_associations_i42_ice <- read_csv('DATA/Eldjarn23_I42_Cardiomyopathy_Significant_Associations_Icelandic.csv')
olink_proteins_dict <- read_csv('DATA/UKB/Eldjarn23_Olink3070_PlasmaProteins.csv')
somascanv4_proteins_dict <- read_csv('DATA/Eldjarn23_SomaScanV4_PlasmaProteins.csv')

ukb_sig_i42 <- eldjarn23_sig_associations_i42_ukb %>%
  select(oid, pval_BI) %>%
  left_join(select(olink_proteins_dict, oid, target_full_name, gene_name), by=c('oid'))

ice_sig_i42 <- eldjarn23_sig_associations_i42_ice %>%
  select(seqid=probe,"pval SMP normalization") %>%
  left_join(select(somascanv4_proteins_dict, seqid, target_full_name, gene_name), by=c('seqid'))

rm(eldjarn23_sig_associations_i42_ukb,eldjarn23_sig_associations_i42_ice, olink_proteins_dict, somascanv4_proteins_dict)
```

These represent the nominally significant associations between plasma proteins in either the UK Biobank or the Icelandic study, with 'Cardiomyopathy'.

## Hypertensive Heart Disease

This is to check for overlap between any HCM patients with hospital diagnoses/death cause record of hypertensive heart disease (I11.0; I11.9).

I then perform the case-control analysis (V3) with non-HTHD HCM patients (independent of HF status) vs. non-HF/non-HCM controls.

```{r}

# hthd_eid <- read_csv('../Plasma_Proteomics/DATA/UKB/RAP/ukb_rap_HTHD_20240209.csv')
# colnames(hthd_eid) <- 'eid'
# 
# sum(ukb_cases$eid %in% hthd_eid$eid) #30 of the UKB cases in all of UKB also have record of hypertensive heart disease
# sum(filter(ukb_pp_hcm_cov, hcm==T)$eid %in% hthd_eid$eid) #3 of the HCM cases in the PP subset also have record of hypertensive heart disease
# sum(filter(ukb_pp_hcm_cov2, hcm==F)$eid %in% hthd_eid$eid) #67 of the non-HCM, non-HF controls in the plasma proteomic subset have hypertensive heart disease but these should be left in the controls
# 
# ukb_pp_hcm_cov3<- ukb_pp_hcm_cov  %>%
#   mutate(hcm = case_when(eid %in% ukb_cases$eid & !eid %in% hthd_eid$eid~ T, #HCM cases independent of HF status
#                          !eid %in% ukb_cases$eid & !eid %in% hf_eid$eid ~ F, #Non-HF non-HCM controls
#                          T~NA)) %>% #non-HCM and HF or HCM and HTHD
#   filter(!is.na(hcm))
# 
# sum(ukb_pp_hcm_cov3$hcm==T)
# 
# limma_hcm_3<- map(olink_cm_panel$ukb_fieldname_equivalent, ~limma_dpe(ukb_pp_hcm_cov3,.)) %>% bind_rows
# write_tsv(limma_hcm_3, './OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/HCM_V3/cmpanel_limma_results.tsv')
# 
# limma_hcm_3 <- left_join(limma_hcm_3, select(olink_cm_panel, Gene, `Protein name`, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent'))
# pp_vs_hcm_3_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(limma_hcm_3, 'allpp_vs_hcm_3',predictor = 'HCM status',output_path='PLOTS/UKB/1_Case_vs_Control/2_Limma/HCM_V3/') 


```

## Power Analysis

This is the power analysis to estimate sample size for single-protein analysis in Jacky's HCM patients.

```{r}
power_analyser <- function(logFCval,sd, power_threshold=0.8, one_or_twosided='one.sided'){
  
  threshold_samplesize <- power.t.test(delta=logFCval,
                                       power=power_threshold,
                                       sd=sd,
                                       sig.level=0.05,
                                       type='two.sample',
                                       alternative=one_or_twosided)
  
  return(threshold_samplesize$n)
}

lnfc <- log(map_dbl(seq(0.2,2,0.2), ~2^.)) #Convert Log2FC to Ln(FC) due to log-normal distribution requiring natural Log
sd <- log(2) #Convert a SD of 1 in Log2FC to a LnFC of 0.69


ss <- map_dbl(lnfc, ~power_analyser(.,sd=sd))
ss <- tibble(log2FC_effect=seq(0.2,2,0.2), ss=ss)

power_plot <- ggplot(ss, aes(x=log2FC_effect, y=ss))+
  geom_point()+
  xlab('Hypothetical Log2FC Effect Size')+
  ylab('Threshold Sample Size in Each Group')+
  labs(title='Power Analysis for Single Protein-Target Analysis via One-Tailed 2-Sample T-test', caption='Assumed Power = 0.8; Significance Threshold=0.05')+
  scale_y_continuous(n.breaks=20)+
  scale_x_continuous(n.breaks=10)

print(power_plot)
```

## Rare-Positive vs. Rare-Negative in UKB-PPP HCM Cases

```{r}
instances <- c('1_I0','2_5Y','3_TOTAL')
casepaths <- str_c('DATA/UKB/PROCESSED/',instances,'/ukb_cmpanelpp_hcm_covariates_bp_t2d_smoking.tsv')
hcm <- map(casepaths, ~read_tsv(.) %>% filter(hcm==T) %>% select(eid))

rare_status <- read_tsv('../../HCM_IntScore/genscore/DATASETS/HCMR_UKB_POSTSPLIT/3_TEST/3_PHENO_WITH_RAREVAR_DOSAGES/allgene/ukb_test_imputedPRS_rareGRSdosage_covariates_cvprs.tsv') %>%
  select(eid=ID, contains('ALL'))

hcm_breakdowns <- map(hcm, ~left_join(., rare_status, by='eid') %>%mutate(rarepos = if_any(contains('ALL'), ~.>=1)))

walk(hcm_breakdowns, ~print(sum(.[['rarepos']], na.rm=T))) #Number of rare positive (across any of the 5 masks) 
walk(hcm_breakdowns, ~print(nrow(.))) #Number of total cases
```

## Undiagnosed Rare+ve HCM Case-Control Analysis

I also run another sensitivity analysis looking at undiagnosed (so non-HCM) individuals who have P/LP variants as per ClinVar (from the GenScore work) in the core 8 genes + FLNC/CSRP3.
These individuals will be taken from the non-HCM/non-HF 'controls' (49,596).

The controls will be the remainder of these individuals who don't have P/LP variants.

Note that `test.tsv` is just to shorten the filename to enable import.

```{r}
# paths <- c('../../HCM_IntScore/genscore/DATASETS/HCMR_UKB_POSTSPLIT/1_TRAIN/3_PHENO_WITH_RAREVAR_DOSAGES/allgene/hcmr_ukb_agesex_train_imputedPRS_rareGRSdosage_covariates_cvprs.tsv',
#            '../../HCM_IntScore/genscore/DATASETS/HCMR_UKB_POSTSPLIT/2_VALIDATION/3_PHENO_WITH_RAREVAR_DOSAGES/allgene/test.tsv',
#            '../../HCM_IntScore/genscore/DATASETS/HCMR_UKB_POSTSPLIT/3_TEST/3_PHENO_WITH_RAREVAR_DOSAGES/allgene/ukb_test_imputedPRS_rareGRSdosage_covariates_cvprs.tsv')
# 
# ukb_rare_dosages <- map(paths, ~read_tsv(.) %>% mutate(ID=as.character(ID)) %>% filter(!str_detect(ID, 'HCR'))) %>% bind_rows()
# 
# ukb_rare_dosages <- mutate(ukb_rare_dosages, rare_pos = ifelse(ALL_clinvar_P_1 >=1 | ALL_clinvar_LP_2 >= 1,T,F))
# 
# ukb_pp_rarepos_hcm <- ukb_pp_hcm_cov  %>%
#   filter(!eid %in% ukb_cases$eid & !eid %in% hf_eid$eid) %>% #Filter for only Non-HF non-HCM individuals
#   mutate(rare = ifelse(eid %in% filter(ukb_rare_dosages, rare_pos==T)$ID, T,F))
# 
# sum(ukb_pp_rarepos_hcm$rare==T)
# sum(ukb_pp_rarepos_hcm$rare==F)
# 
# limma_hcm_rarepos<- map(olink_cm_panel$ukb_fieldname_equivalent, ~limma_dpe(ukb_pp_rarepos_hcm,., predictor_vars = 'rare')) %>% bind_rows
# write_tsv(limma_hcm_rarepos, './OUTPUT/UKB/PLOTS/1_Case_vs_Control/2_Limma/HCM_RarePos/cmpanel_limma_results.tsv')
# 
# limma_hcm_rarepos <- left_join(limma_hcm_rarepos, select(olink_cm_panel, Gene, `Protein name`, ukb_fieldname_equivalent), by=c('pp'='ukb_fieldname_equivalent'))
# pp_vs_hcm_rarepos_marginal_bonferroni_summarytb_limma <- summary_plotter_limma(limma_hcm_rarepos, 'allpp_vs_hcm_rarepos',predictor = 'Rare status',output_path='PLOTS/UKB/1_Case_vs_Control/2_Limma/HCM_RarePos/') 
```

##Pairwise-Correlation Analysis with NTproBNP

This evaluates multicollinearity via evaluation of all biomarker's pairwise correlation with NTproBNP (prior to transformation btw).

```{r}
pp <- colnames(ukb_proteomics_data)[-1]

pearson_pairwise_ntprobnp<- map(pp, ~cor(ukb_proteomics_data[[.]], ukb_proteomics_data$ntprobnp, use='pairwise.complete.obs')) 
names(pearson_pairwise_ntprobnp) <- pp

pearson_pairwise_ntprobnp <- pearson_pairwise_ntprobnp %>%
  bind_rows() %>%
  pivot_longer(everything(), names_to='pp', values_to='r')
```


## PRS Addition to ML Dataset

```{r}
#Filter out those with NA values in cv_prs
# ukb_pp <- ukb_pp %>%
#   filter(!is.na(cv_prs))
# 
# sum(ukb_pp$hcm)
# sum(ukb_pp$hcm ==F)

# write_tsv(ukb_pp,str_c('DATA/UKB/PROCESSED/',instance,'/ukb_ALLpanel_pp_hcm_covariates_bp_t2d_smoking_rcmmcov_cvprs.tsv'))

# rm(ukb_pp, ukb_cv_prs)
```
In reality, to maintain the same training and test sets are when there is no PRS added, I take the X_train and X_test datasets which have previously been split and add the cv_prs from the above tibble onto them and I rewrite it out.

```{r}

prs_adder <- function(prs_tb, folder_path='OUTPUT/UKB/ML/1_data/1_hcm_cc_noprs/'){
  
  files <- list.files(folder_path, full.names=T)
  datasets <- map(files, ~read_csv(.) %>% mutate(index=row_number()))
  names(datasets) <- str_match(files, str_c(folder_path,'([^\\.]+)\\.csv'))[,2]
  
  #Add the prs to each of the two datasets but can't do so by ID because ID lost in X_train and X_test so instead need to align the 10 genetic PCs
  
  X_datasets <- c(datasets['X_train'],datasets['X_test'])
  y_datasets <- c(datasets['y_train'],datasets['y_test'])
  rm(datasets)
  
  #Rename the prs_tb PC columns
  colnames(prs_tb) <- c('ID','cv_prs', str_c('pc',1:10))
  
  joined_tbs <- map(X_datasets, ~left_join(., prs_tb, by=str_c('pc',1:10)))
  
  
}


prs_X_train <- hcm_cc_5y_X_train %>%
  left_join(ukb_cv_prs, by = )

#Filter for those without cv_prs

```